---
title: "Ch21_Iteration"
format: html
editor: visual
---

## 

# Introduction

Reducing code duplication has three main benefits:

1.  easier to read the intent of the code
2.  easier to make coding changes (i.e. pivot)
3.  fewer bugs or typos

#### Q: Based on your previous lecture, what approach allowed to reduce code? 

Here we will introduce a second way of reducing duplication: iteration

iteration —\> helps you when you need to do the same thing to multiple inputs...

```{r}
library(tidyverse)

```

## 21.2 For loops

```{r}

df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)
```

If we want to compute median for each columns

```{r}

median(df$a)
median(df$b)
median(df$c)
median(df$d)

```

#### Replace code above with for loop:

```{r}
output <- vector("double", ncol(df))  # 1. output
output # defines a empty vector 

for (i in seq_along(df)) {            # 2. sequence
  output[[i]] <- median(df[[i]])      # 3. body
}

output
```

EEveryone for loop has 3 components:

1.  The output: `output <- vector("double", length(x))`. Allocate space for the output... This speeds up the for loop.
2.  The sequence: `i in seq_along(df)` . Determins what to loop over: each run of the for loop will assign i to different values from `seq_along(df)`

-   note: `seq_along()` similar to `1:length(l)` where `l` is the length of the vector or column.

Example:

```{r}

y <- vector("double", 4)
seq_along(y)

1:length(y)
```

3.  The body: `output[[i]] <- median(df[[i]])` . This is the main part of the for loop code!

#### 21.2.1 Exercises

1.  Write for loops to:

    1.  Compute the mean of every column in `mtcars`.

    2.  Determine the type of each column in [`nycflights13::flights`](https://rdrr.io/pkg/nycflights13/man/flights.html).

    3.  Compute the number of unique values in each column of `iris`.

    4.  Generate 10 random normals from distributions with means of -10, 0, 10, and 100.

    Think about the output, sequence, and body **before** you start writing the loop.

```{r}

```

## 21. 3 For loop variations

Different variations of `for loop` .

Four variations of the basic them of the for loop:

1.  Modifying an exiting object, instead of creating a new object
2.  Looping over names or values, instead of indices
3.  Handling outputs of unknown length
4.  Handling sequences of unknown length

#### 21.3.1 Modifying an existing object

```{r}
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)
rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}

df$a <- rescale01(df$a)
df$b <- rescale01(df$b)
df$c <- rescale01(df$c)
df$d <- rescale01(df$d)
```

How can we solve this with a for loop? Think of this 3 components:

1.) Output: swap the dataframe column, so it is the same as input

2.) Sequence: iterate over each column using `seq_along()`

3.) Body: apply `rescale01()`

```{r}

# DO NOT RUN MORE THAN ONCE
for (i in seq_along(df)) {
  df[[i]] <- rescale01(df[[i]]) # swap column
}
```

### 21.3.2 Looping patterns

looping over the numeric indices with `for (i in seq_along(xs))`, and extracting the value with `x[[i]]`.

1.  Loop over the elements: `for (x in xs))`.
2.  Loop over the names: `for (nm in names(xs))` .

Q: what is the advantage of 2.) over 1.)? Do you have an example?

#### note: numeric indices are more general (see example)

```{r}

results <- vector("list", length(x))
names(results) <- names(x)

for (i in seq_along(x)) {
  name <- names(x)[[i]]
  value <- x[[i]]
  
}
```

### 21.3.3 Unknown output length

What happens when you don't know the length of the vector?

we can progressively grow the vector

```{r}
# create a list
means <- c(0, 1, 2)

# step 0 
output <- double()

# step 1
for (i in seq_along(means)) {
  # step 2
  n <- sample(100, 1)
  output <- c(output, rnorm(n, means[[i]]))
}
str(output)
```

#### Q: What is the problem with this approach? 

R has to copy the data from the previous iteration. This means that there is a quadratic time complexity O(n\^2). AVOID!

```{r}


out <- vector("list", length(means))
for (i in seq_along(means)) {
  n <- sample(100, 1)
  out[[i]] <- rnorm(n, means[[i]])
}

# list of vectors
str(out)

# flatten a list of vectors into a single vector. 
str(unlist(out))

```

### 21.3.4 Unknown sequence length

We want to stop the loop when a condition is meant... E.g., three heads were achieved.

Use the while loop

```{r}

while (condition) {
  # body
}
```

#### What is the relationship between for and while loop? 

```{r}

for (i in seq_along(x)) {
  # body
}

# Equivalent to
i <- 1
while (i <= length(x)) {
  # body
 
 i <- i + 1 
}
```

#### Example: 

```{r}

flip <- function() sample(c("T", "H"), 1)

flips <- 0
nheads <- 0

while (nheads < 3) {
  if (flip() == "H") {
    nheads <- nheads + 1
  } else {
    nheads <- 0
  }
  flips <- flips + 1
}
flips
```

## 21.4 For loops vs functionals

For loops are not as important in R versus other languages (e.g. python) since R is a functional programming language.

```{r}

df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)

```

let's compute the mean of every column. Use the for loop like before

```{r}
output <- vector("double", length(df))
for (i in seq_along(df)) {
  output[[i]] <- mean(df[[i]])
}
output

```

functional programming example: write the for loop in a function.

```{r}

col_mean <- function(df) {
  output <- vector("double", length(df))
  for (i in seq_along(df)) {
    output[i] <- mean(df[[i]])
  }
  output
}


# run function
col_mean(df)
```

Let's also compute the median, standard deviation...

```{r}

col_median <- function(df) {
  output <- vector("double", length(df))
  for (i in seq_along(df)) {
    output[i] <- median(df[[i]])
  }
  output
}
col_sd <- function(df) {
  output <- vector("double", length(df))
  for (i in seq_along(df)) {
    output[i] <- sd(df[[i]])
  }
  output
}

col_median(df)
col_sd(df)


```

#### Exercise: can you rewrite these functions into 1?

```{r}

f1 <- function(x) abs(x - mean(x)) ^ 1
f2 <- function(x) abs(x - mean(x)) ^ 2
f3 <- function(x) abs(x - mean(x)) ^ 3
```

===========================

\< Ans\>

===========================

Instead of making `col_mean()`,`col_median()`, `col_sd()` by adding an `fun` argument.

```{r}

col_summary <- function(df, fun) {
  out <- vector("double", length(df))
  for (i in seq_along(df)) {
    out[i] <- fun(df[[i]])
  }
  out
}
col_summary(df, median)
col_summary(df, mean)
col_summary(df, sd)

```
